Placeholder for article 3

### 2025-03-01 analysis/paper/paper.qmd
Macbook daily work session. Results in main document corrected and rewritten for emphasis on main patterns and comparison between groups. Supplementary document revised and corrected.

After the homogenization and merging of the datasets, descriptive statistical analysis was applied to describe and interpret the relation between the raw material and technological data. As mentioned above, the goal of the study was to test two hypotheses related to the technological organisation of local and non-local cherts, and between different UP occupations. To test these hypotheses, we evaluated several parameters for each dataset and assemblage and compared the results between chert types and assemblages. The parameters evaluated in the current study were: P1) frequency of retouched tools; P2) frequency of maintenance products; P3) debitage sizes; P4) tools-to-debitage ratios; P5) core sizes; P6) blank-to-core ratios; P7) retouched tool sizes; P8) tool diversity; P9) tool specialization. Debitage, core and retouched tool sizes were analysed through descriptive statistics (mean, sd, min and max, tables available in the SOM\*\*) and plotted through boxplots. The ratios were obtained through the division of each group (e.g., tools by debitage and blanks by core), by chert type and assemblage. All other parameters were observed through number and percentages tables. All the analyses were done through code in R Studio.

@@

In all assemblages, blanks and blank fragments are the most frequent class of lithic artefacts and in all chert types (>40% of the assemblages). Cores are, as expected, found in smaller percentages in all assemblages (~3-8%). Cores are present in local cherts in larger numbers than in non-local types (P++); their presence is noted in almost all non-local cherts in all assemblages, except for the TL group. In the Proto-Solutrean assemblage however, non-local chert core presence is small and limited to T7-9 (n=4 complete cores and n=2 core fragments). Albeit in very small numbers in all assemblages (n<10), core preparation and maintenance products (P2) are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7. The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian and Proto-Solutrean assemblages. Retouched tools (P1) were identified in all chert types from all studied assemblages; in general, all assemblages show a similar pattern where local cherts show smaller percentages of retouched tools than non-local cherts. However, the assemblages show clear differences in the number of retouched tools.

@@

The relation between blank and core numbers was explored through a blank-to-core ratio (P6; Figure 1 and table s++). In all assemblages, local cherts have generally smaller ratios than non-local cherts, although sample sizes for cores are variable (graph ++). For the Gravettian assemblages, the results show that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4. The Solutrean assemblage also shows higher ratios for non-local cherts (>12), with local cherts showing ratios below 7.  The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7. However, non-local cores are poorly represented within the assemblage (table ++).

@@

Retouched tools were further explored through the tools-to-debitage ratio (P4; Figure 2 and table s++). In general, the result show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages. The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5. In the Proto-Solutrean non-local cherts show ratios between 0.3 to 0.5, while local cherts show ratios below 0.3. This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages. Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of the local chalcedony).

@@

The sizes of blanks (P3) and cores (P5) are explored through scatterplots of width and length (Figure 3 a) and boxplots for weight (Figure 3 b); complete tables with descriptive statistics and non-parametric statistical test results can also be found in SOM ++, table s++. The scatterplots show different patterns in the several analysed classes.

@@

ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")

@@

ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")

@@

ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a", "b"), ncol = 2, common.legend = TRUE, legend = "bottom")

@@

In the Gravettian assemblage, non-local chert blanks (P3) are smaller than local chert blanks, especially regarding length (~ <25 mm) and weight (<5 gr), with statistically significant differences (p value of 0.00009*** and 0.012, respectively). The local chert blanks, although mostly concentrated below 25 mm width and length and with weights below 5 gr, have the largest number of outliers and variability in terms of size. The same patterns can be seen in the Proto-Solutrean assemblage, where length and weight also show statistically significant differences between chert types (0.046* and 0.004), with non-local types showing smaller values. In comparison, in the Solutrean assemblage, only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005**) than local chert blanks, being mostly concentrated below ~30 mm, while local chert blank lengths are more variable. While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02).

Regarding cores (P5), the differences in sizes and weights seem between local and non-local cherts seem to be less obvious than those observed in blanks. In the Gravettian assemblage, there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable. Especially regarding weight, T6 cores (non-local) show lighter cores (~12 gr means) compared to local cherts. In the Solutrean assemblage, weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (~18 gr means) than local cherts (~22 gr means). Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p > 0.05). The Proto-Solutrean assemblage shows a small sample of non-local chert cores (n=4), not allowing a clear identification of patterns between the groups; alike the other assemblage however, local chert cores show a high range of sizes (~15-50 mm width and length) and weights (~3-85 gr).

@@

In the Gravettian assemblage, retouched tools (P1) are present in higher frequencies in the non-local cherts, as well as in core preparation and maintenance products (P2). Debitage sizes (P3) show smaller sizes and weights for non-local cherts. Tools-to-debitage ratios (P4) are significantly higher in non-local cherts. Core sizes (P5) are slightly smaller in T6, a non-local chert, and the blank-to-core ratios (P6) are higher in non-local cherts. Retouched sizes (P7) show no relevant differences between local and non-local cherts; however, tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9). The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.

The Proto-Solutrean assemblages show a smaller presence of retouched tools than the Gravettian, and within the assemblage, retouched tools (P1) are slightly more frequent in non-local cherts. Core preparation and maintenance products (P2) are only present in non-local cherts, albeit in very small numbers (n=2). Blank sizes (P3), core sizes (P5), and retouched tool sizes (P7) are slightly smaller in non-local cherts. Regarding the ratios, tool-to-debitage ratios (P4) are higher in non-local cherts, albeit smaller than those seen in the Gravettian assemblage, and blank-to-core ratios (P6) are also higher in non-local cherts. Little differences were observed in tool diversity (P8) and specialisation (P9) between local and non-local cherts.

Finally, the Solutrean shows no relevant differences in the frequencies of retouched tools (P1) and core preparation and maintenance products (P2) between local and non-local cherts, with only slightly higher frequencies in the non-local cherts. In general, the retouch frequency is lower than that observed in the previous assemblages. Debitage (P3), core (P5), and retouched tool (P7) sizes show no differences between local and non-local cherts, except for elongated blanks which are smaller in T6 and bigger in T2, and for slightly smaller cores seen in T6, although with high standard deviations. Similarly to the Proto-Solutrean, tools-to-debitage ratios (P4) and blank-to-core ratios (P6) are slightly higher in non-local cherts. Both tool diversity (P8) and tool specialization (P9) are higher in local cherts.

>>>>

After the homogenization and merging of the datasets, descriptive statistical analysis was applied to describe and interpret the relation between the raw material and technological data. As mentioned above, the goal of the study was to test two hypotheses related to the technological organisation of local and non-local cherts, and between different UP occupations. To test these hypotheses, we evaluated several parameters for each dataset and assemblage and compared the results between chert types and assemblages. The parameters evaluated in the current study were: P1) frequency of retouched tools; P2) frequency of cores; P3) frequency of maintenance products; P4) debitage sizes; P5) tools-to-debitage ratios; P6) core sizes; P7) blank-to-core ratios; P8) tool diversity; P9) tool specialization. Debitage and core sizes were analysed through descriptive statistics (mean and sd tables available in the SOM\*\*) and plotted through boxplots and through non-parametric statistical analyses (Mann-Whitney U test). The ratios were obtained through the division of each group (e.g., tools by debitage and blanks by core), by chert type and assemblage. All other parameters were observed through number and percentages tables. All the analyses were done through code in R Studio.

@@

The results are presented following the analysis parameters, focusing on the comparison between wider chert groups (local and non-local cherts) and between assemblages. A complete description of the assemblages and detailed tables for each analysis presented in this study can be found in the supplementary materials (SOM ++).

In all assemblages, blanks and blank fragments are the most frequent class of lithic artefacts and in all chert types (>40% of the assemblages). Cores are, as expected, found in smaller percentages in all assemblages (~3-8%). Cores are present in local cherts in larger numbers than in non-local types (P2); their presence is noted in almost all non-local cherts in all assemblages, except for the TL group. In the Proto-Solutrean assemblage however, non-local chert core presence is small and limited to T7-9 (n=4 complete cores and n=2 core fragments). Albeit in very small numbers in all assemblages (n<10), core preparation and maintenance products (P3) are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7. The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian and Proto-Solutrean assemblages. Retouched tools (P1) were identified in all chert types from all studied assemblages; in general, all assemblages show a similar pattern where local cherts show smaller percentages of retouched tools than non-local cherts. However, the assemblages show clear differences in the number of retouched tools.

@@

The relation between blank and core numbers was explored through a blank-to-core ratio (P7; Figure 1 and table s++). In all assemblages, local cherts have generally smaller ratios than non-local cherts, although sample sizes for cores are variable (graph ++). For the Gravettian assemblages, the results show that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4. The Solutrean assemblage also shows higher ratios for non-local cherts (>12), with local cherts showing ratios below 7.  The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7. However, non-local cores are poorly represented within the assemblage (table ++).

@@

Retouched tools were further explored through the tools-to-debitage ratio (P5; Figure 2 and table s++). In general, the result show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages. The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5. In the Proto-Solutrean non-local cherts show ratios between 0.3 to 0.5, while local cherts show ratios below 0.3. This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages. Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of the local chalcedony).

@@

The sizes of blanks (P4) and cores (P6) are explored through scatterplots of width and length (Figure 3 a) and boxplots for weight (Figure 3 b); complete tables with descriptive statistics and non-parametric statistical test results can also be found in SOM ++, table s++. The scatterplots show different patterns in the several analysed classes.

@@

ggpubr::ggarrange(gravettian_scatter, gravettian_stats_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")

@@

ggpubr::ggarrange(proto_scatter, proto_stats_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")

@@

ggpubr::ggarrange(solutrean_scatter, solutrean_stats_weight, labels = c("a", "b"), nrow = 2, common.legend = TRUE, legend = "bottom")

@@

In the Gravettian assemblage, non-local chert blanks (P4) are smaller than local chert blanks, especially regarding length (~ <25 mm) and weight (<5 gr), with statistically significant differences (p value of 0.00009*** and 0.012, respectively). The local chert blanks, although mostly concentrated below 25 mm width and length and with weights below 5 gr, have the largest number of outliers and variability in terms of size. The same patterns can be seen in the Proto-Solutrean assemblage, where length and weight also show statistically significant differences between chert types (0.046* and 0.004), with non-local types showing smaller values. In comparison, in the Solutrean assemblage, only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005**) than local chert blanks, being mostly concentrated below ~30 mm, while local chert blank lengths are more variable. While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02).

Regarding cores (P6), the differences in sizes and weights seem between local and non-local cherts seem to be less obvious than those observed in blanks. In the Gravettian assemblage, there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable. Especially regarding weight, T6 cores (non-local) show lighter cores (~12 gr means) compared to local cherts. In the Solutrean assemblage, weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (~18 gr means) than local cherts (~22 gr means). Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p > 0.05). The Proto-Solutrean assemblage shows a small sample of non-local chert cores (n=4), not allowing a clear identification of patterns between the groups; alike the other assemblage however, local chert cores show a high range of sizes (~15-50 mm width and length) and weights (~3-85 gr).

@@

In the Gravettian assemblage, retouched tools (P1) are present in higher frequencies in the non-local cherts, as well as in core preparation and maintenance products (P3). Cores are present in even in non-local chert types (P2). Debitage sizes (P4) show smaller sizes and weights for non-local cherts. Tools-to-debitage ratios (P5) are significantly higher in non-local cherts. Core sizes (P6) are slightly smaller in T6, a non-local chert, and the blank-to-core ratios (P7) are higher in non-local cherts. Tool diversity (P8) is higher in non-local cherts, although with high frequency of less specialised tools such as notches and retouched tools, alongside hunting implements such as backed bladelets (P9). The high frequency of burins seen in this assemblage may be related to their use as cores, representing a knapping strategy more than the presence of retouch.

The Proto-Solutrean assemblages show a smaller presence of retouched tools than the Gravettian, and within the assemblage, retouched tools (P1) are slightly more frequent in non-local cherts. Cores of non-local chert types are found in very small numbers (P2), although knapping is likely occurring on site. Core preparation and maintenance products (P3) are only present in non-local cherts, albeit in very small numbers (n=2). Blank sizes (P4) and core sizes (P6) are slightly smaller in non-local cherts. Regarding the ratios, tool-to-debitage ratios (P5) are higher in non-local cherts, albeit smaller than those seen in the Gravettian assemblage, and blank-to-core ratios (P7) are also higher in non-local cherts. Little differences were observed in tool diversity (P8) and specialisation (P9) between local and non-local cherts.

Finally, the Solutrean shows no relevant differences in the frequencies of retouched tools (P1) and core preparation and maintenance products (P2) between local and non-local cherts, with only slightly higher frequencies in the non-local cherts. In general, the retouch frequency is lower than that observed in the previous assemblages. Cores are present in relevant quantities in non-local chert types (P2). Debitage (P4) mass is slightly smaller in non-local cherts, although with larger widths. Core (P6) sizes show slightly smaller values for T6. Similarly to the Proto-Solutrean, tools-to-debitage ratios (P5) and blank-to-core ratios (P7) are slightly higher in non-local cherts. Both tool diversity (P8) and tool specialization (P9) are higher in local cherts.
